On Devon Allen's Disqualification at the 2022 Track and Field World Championship

### 2023-04-20 Manuscript/manuscript.tex
Updated tables, added some stuff based of Tuesday meeting notes

\subsection{Beyond World Championship}
This section was motivated from a qualitative analysis that started by looking
at how United States athletes reacted at the 2022 United States Track and Field
Championships. From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teammates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
\jy{cross-reference to the table by its label}
too quickly.  Table~\ref{fig:USAvsWorld} below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. ``USA" denotes
the USA Track and Field Championships, ``World" denotes the World Track and Field
Championships, ``N/A" was used for any athlete that did not compete and thus
did not register a reaction time.  All numbers listed in the table are the reaction
time for the respective athletes in seconds.

\begin{center}
  \begin{tabular}{c c c c c c c} 
   \toprule
   Athlete & USA H & USA S & USA F & World H & World S & World F \\ [0.5ex] 
   \midrule
   Devon Allen & 0.201 & 0.153 & 0.160 & 0.123 & 0.101 & 0.099 \\ 
   Trey Cunningham & 0.186 & 0.185 & 0.182 & 0.115 & 0.120 & 0.109 \\
   Grant Holloway & 0.192 & 0.190 & N/A & 0.147 & 0.128 & 0.124 \\
   Daniel Roberts & 0.181 & 0.200 & 0.183 & 0.179 & N/A & N/A \\ [0.5ex]
   \bottomrule
   \label{fig:USAvsWorld}
  \end{tabular}
  \end{center}

@@

\jy{Undefined notations}
The test statistic of the DS method of the Wilcoxon rank-sum statistic is:
$Z = \frac{S - E(S)}{\sqrt{\hat{\text{Var}}(S)}}$ where 
$W^* = \frac{1}{N} + \sum_{i=1}^{N} \delta_{i}^{*} R_{i}^{*}$, $R_{i}^{*}$ is
the rank of $X_{i}^{*}$.
The asymptomatic distribution of the Wilcoxon rank-sum statistic was found by
\citet{datta2005rank} and the R package ``clusrank" allows for its implementation.
\citep{jiang2017wilcoxon}.

Using the \texttt{clusWilcox.test()} function, a Wilcoxon rank sum
test was performed with the cluster set to be the Athlete and group set to Year
using the DS method. The DS method is preferable when the cluster sizes are
informative \citep{jiang2017wilcoxon}. \jy{explain. More data means faster
  reaction time?}
We do not care that some
athletes made it further in either competition than another or that it should
not matter how elite an athlete is at the 110 meter hurdles once we get to the
national international competition level.  Cluster sizes vary from 2 observations
to 6.

@@

As shown in table~\ref{fig:Gamma_parameters}, the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022.  The men's final data venue effect error increased from 0.439
to 0.504 and from 0.0495 to 0.0554 for the men's pooled data.  This demonstrates
that the inclusion of 2022 increased the randomness of the data and motivates
the rank-based study~\ref{subsec:Results_Rank}.

  Gamma model with venue effect and link=log
  \begin{center}
    \begin{tabular}{c  c  c  c  c  c  c } 
     \toprule
     Dataset & Venue standard error & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
     \midrule
     Older Finals & 0.0439 & -1.901 & 0.0251 & 0.0019 & 0.0109 \\
     All Finals & 0.0504 & -1.914 & 0.0280 & 0.0025 & 0.0112 \\
     Older Pooled & 0.0495 & -1.853 & 0.0305 & 0.0024 & 0.0243 \\
     All Pooled & 0.0554 & -1.869 & 0.0348 & 0.0031 & 0.0235 \\
     \bottomrule
     \label{fig:Gamma_parameters}
    \end{tabular}
    \end{center}

At a size of $n=10,000,000$, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:
\jy{label the table and reference to it; clean it up}
\begin{center}
  \begin{tabular}{c c} 
   \toprule
   Data Set & Probability RxnTime $<$ 0.1 \\ 
   \midrule
   all finals gamma log & 0.000856 \\ 
   old finals gamma log & 0.000439 \\
   all pooled gamma log & 0.006865 \\
   old pooled gamma log & 0.005761 \\
   \bottomrule
   \label{fig:Sim_probability}
  \end{tabular}
  \end{center}

Using the gamma function with link set to log, the probability of the function being
less than 0.1 seconds can be calculated to see if Allen's disqualification was an anomaly.
Prior to the 2022 World Championship, there was a 0.0439\% chance (based on the
simulated results) of observing a reaction time below 0.01 seconds in the men's
finals as shown in table~\ref{fig:Sim_probability}.
That means that we would expect a time below the threshold to occur
once every 2,277 starts.  Considering there are roughly 8 men's finals reaction
times every 2 years (The World Championship is biennial), every 569 years there
should be a time as extreme as Allen's.  This seems extremely unlikely, and thus
it seems like there may be an additional explanation for the fast reaction times.

@@

It is worth noting that changing the reaction time barrier from 0.1 seconds to
0.08 seconds drastically reduces the chances of observing a reaction time that
would break the barrier.  For the dataset of all finals data, the probability
associated with observing a reaction time below 0.08 seconds was $6\cdot10^{-7}$.
If IAAF were to change the reaction time the way that \citet{komi2009iaaf}'s paper
suggests, the probability would be so much lower that athletes would likely not
have a case to make when arguing against a disqualification.

>>>>

\subsection{Beyond World Championship}
This section was motivated from a qualitative analysis that started by looking
at how United States athletes reacted at the 2022 United States Track and Field
Championships. From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teammates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
\jy{cross-reference to the table by its label}
too quickly.  Table~\ref{fig:USAvsWorld} below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. ``USA" denotes
the USA Track and Field Championships, ``World" denotes the World Track and Field
Championships,   All numbers listed in the table are the reaction
time for the respective athletes in seconds.

\begin{table}
\begin{center}
  \begin{tabular}{c c c c c c c} 
   \toprule
   Athlete & USA H & USA S & USA F & World H & World S & World F \\ [0.5ex] 
   \midrule
   Devon Allen & 0.201 & 0.153 & 0.160 & 0.123 & 0.101 & 0.099 \\ 
   Trey Cunningham & 0.186 & 0.185 & 0.182 & 0.115 & 0.120 & 0.109 \\
   Grant Holloway & 0.192 & 0.190 & N/A & 0.147 & 0.128 & 0.124 \\
   Daniel Roberts & 0.181 & 0.200 & 0.183 & 0.179 & N/A & N/A \\ [0.5ex]
   \bottomrule
  \end{tabular}
  \label{fig:USAvsWorld}
  \caption{Rection times for four United States Athletes at the 2022 USA Track
  and Field World Championships and the 2022 World Track and Field Championship.
  ``N/A" was used for any athlete that did not compete and thus did not register 
  a reaction time.}
  \end{center}
\end{table}

@@

\jy{Undefined notations}
The test statistic of the DS method of the Wilcoxon rank-sum statistic is:
$Z = \frac{S - E(S)}{\sqrt{\hat{\text{Var}}(S)}}$ where 
$W^* = \frac{1}{N} + \sum_{i=1}^{N} \delta_{i}^{*} R_{i}^{*}$, $R_{i}^{*}$ is
the rank of $X_{i}^{*}$. S is given as $S = \mathbb{E}[W^* | X, \delta]$ and
$E(S) = E(W^*) = \frac{1}{2} \sum_{i=1}^{N} \frac{n_i}{N}$.
The asymptomatic distribution of the Wilcoxon rank-sum statistic was found by
\citet{datta2005rank} and the R package ``clusrank" allows for its implementation.
\citep{jiang2017wilcoxon}.
---Please see manuscript around line 464 for comments regarding this section---

%I included additional statistics from the paper, I am unsure of how in depth
%This explanation needs to be

%$F_{i}(x) = \frac{1}{n_i} \sum_{k=1}^{n_i} \mathbb{I}(X_{i,k} \leq x)$

%$F_{i}(x-) = \frac{1}{n_i} \sum_{k=1}^{n_i} \mathbb{I}(X_{i,k} < x)$

%$S = \frac{1}{N + 1} \sum_{i=1}^{N}  \sum_{k=1}^{n_i} \frac{\delta_{ik}}{n_i} \left[ 1 + \frac{1}{2} \sum_{j \neq i} \frac{1}{n_j}
%\left( F_j(X_{ik}) + F_j(X_{ik-}) \right) \right]$

Using the \texttt{clusWilcox.test()} function, a Wilcoxon rank sum
test was performed with the cluster set to be the Athlete and group set to Year
using the DS method. The DS method is preferable when the cluster sizes are
informative \citep{jiang2017wilcoxon}.  In this case, athletes who have more
data are generally thought to be faster, as they have made it further in the
competition.  Clusters range from two to six, with two indicating that at each
competition they only made it to the heats, but a cluster size of six indicates
that they made the finals in both competitions.  While the number of times
athletes compete is based off their overall time not their reaction time, the
two are correlated \citep{delalija2008reaction}.

@@

As shown in table~\ref{tab:Gamma_parameters}, the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022.  The men's final data venue effect error increased from 0.439
to 0.504 and from 0.0495 to 0.0554 for the men's pooled data.  This demonstrates
that the inclusion of 2022 increased the randomness of the data and motivates
the rank-based study~\ref{subsec:Results_Rank}.

\begin{table}
  \centering
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & Venue standard error & Int effect est & Int se & Int var & Resid Var \\
      \midrule
      Older Finals & 0.0439 & $-1.901$ & 0.0251 & 0.0019 & 0.0109 \\
      All Finals & 0.0504 & $-1.914$ & 0.0280 & 0.0025 & 0.0112 \\
      Older Pooled & 0.0495 & $-1.853$ & 0.0305 & 0.0024 & 0.0243 \\
      All Pooled & 0.0554 & $-1.869$ & 0.0348 & 0.0031 & 0.0235 \\
      \bottomrule
  \end{tabular}
  \caption{Summary statistics for the GLMM on various data sets}
  \label{tab:Gamma_parameters}
\end{table}

At a size of $n=10,000,000$, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:
\jy{label the table and reference to it; clean it up}
\begin{table}
  \centering
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & P(RxnTime $<$ 0.1) & P(RxnTime $<$ 0.09) & P(RxnTime $<$ 0.08) \\ 
   \midrule
   all finals gamma log & $8.56\cdot10^{-4}$ & $4.1\cdot10^{-5}$ & $6.0\cdot10^{-7}$ \\ 
   old finals gamma log & $4.39\cdot10^{-4}$ & $1.63\cdot10^{-5}$ & $2.0\cdot10^{-5}$ \\
   all pooled gamma log & $6.86\cdot10^{-3}$ & $1.21\cdot10^{-3}$ & $1.3\cdot10^{-4}$ \\
   old pooled gamma log & $5.76\cdot10^{-3}$ & $9.93\cdot10^{-4}$ & $1.0\cdot10^{-4}$\\
   \bottomrule
  \end{tabular}
  \caption{Probabilities of observing reaction times less than 0.1, 0.09, and
  0.08 for various data sets}
  \label{tab:Sim_probability}
\end{table}

Using the gamma function with link set to log, the probability of the function being
less than 0.1 seconds can be calculated to see if Allen's disqualification was an anomaly.
Prior to the 2022 World Championship, there was a 0.0439\% chance (based on the
simulated results) of observing a reaction time below 0.01 seconds in the men's
finals as shown in table~\ref{tab:Sim_probability}.
That means that we would expect a time below the threshold to occur
once every 2,277 starts.  Considering there are roughly 8 men's finals reaction
times every 2 years (The World Championship is biennial), every 569 years there
should be a time as extreme as Allen's.  This seems extremely unlikely, and thus
it seems like there may be an additional explanation for the fast reaction times.

@@

Table ~\ref{tab:Sim_probability} shows that changing the reaction time barrier from 0.1 seconds to
0.08 seconds drastically reduces the chances of observing a reaction time that
would break the barrier.  For the dataset of all finals data, the probability
associated with observing a reaction time below 0.08 seconds was $6\cdot10^{-7}$.
If IAAF were to change the reaction time the way that \citet{komi2009iaaf}'s paper
suggests, the probability would be so much lower that athletes would likely not
have a case to make when arguing against a disqualification.
